\startcomponent curriculum-vitae

\environment layout

\startBackmatterHead[title={Curriculum Vitae}]

Gydo Cornelis Petrus van Zundert was born in Oud Gastel on the 12th of June, 1987. He
attended the local Sint Bernardus School, where he graduated with the maximum Cito
score of 550 and co-starred in the final school play. After an uninspiring yet
fun time at the VWO of the Sint Norbertus College in Roosendaal, Gydo started
his Bachelor \emph{Chemistry} in 2005 at Utrecht University. He performed his
Bachelor thesis in the Theoretical Chemistry group of \bold{dr. Joop van
Lenthe}. Next, he registered for the Master \emph{Nanomaterials: chemistry and
physics}, and did his Master's internship in the MolDyn group at the FOM
Institute for Plasma Physics Rijnhuizen, in Nieuwegein, under the tutelage of
\bold{dr. Anouk Rijs}. His work resulted in a first-author paper and he
graduated with the maximum GPA of 4.0.

On the 1st of November 2011 he began his doctorate in the Computational
Structural Biology group of \bold{prof. dr.  Alexandre Bonvin}. During these
years he received a poster price (combined ACMM/NSBM symposium) and two travel
grants (Keystone Symposia Future of Science Scholarship, and Instruct Biennial
Meeting Scholarship), and he was the PhD representative of the NMR department
in 2014 and 2015. His scientific work has been published in high-quality
peer-reviewed journals. Gydo will defend his dissertation on the 25th of
November 2015.

\stopBackmatterHead

\stopcomponent
