\startcomponent publications

\environment layout

\startBackmatterHead[title=Publications]

{\bf In this thesis:}

\startitemize

\item {\bf G.C.P van Zundert}, J.P.G.L.M. Rodrigues, M. Trellet, C. Schmitz, P.L.
Kastritis, E. Karaca, A.S.J. Melquiond, M. van Dijk, S.J. de Vries and A.M.J.J.
Bonvin. The HADDOCK2.2 web server: User-friendly integrative modeling of
biomolecular complexes. J. Mol. Biol. In press (2015).

\item {\bf G.C.P. van Zundert} and A.M.J.J. Bonvin. DisVis: quantifying and visualizing
accessible interaction space of distance-restrained biomolecular complexes.
Bioinformatics, Advanced Online Publication (2015).

\item {\bf G.C.P. van Zundert}, A.S.J. Melquiond and A.M.J.J. Bonvin.
Integrative modeling of biomolecular complexes: HADDOCKing with cryo-electron
microscopy data. Structure 23, 949--960 (2015).

\item {\bf G.C.P. van Zundert} and A.M.J.J. Bonvin. Fast and
sensitive rigid-body fitting into cryo-EM density maps with PowerFit. AIMS
Biophysics, 2, 73--87 (2015).

\stopitemize

{\bf Others:}

\startitemize

\item M. Kaplan, A. Cukkemane, {\bf G.C.P. van Zundert}, S. Narasimhan, M.
Daniëls, D. Mance, G. Waksman, A.M.J.J. Bonvin, R. Fronzes, G.E. Folkers and M.
Baldus. Probing a cell-embedded megadalton protein complex by DNP-supported
soid-state NMR. Nat. Methods 12, 649--652 (2015).

\item {\bf G.C.P. van Zundert} and A.M.J.J. Bonvin. Modelling protein-protein
complexes using the HADDOCK webserver. Methods Mol. Biol. 1137, 163--179 (2014).

\item {J.P.G.L.M. Rodrigues, A.S.J. Melquiond, E. Karaca, M. Trellet, M. Van
Dijk, {\bf G.C.P. Van Zundert}, C. Schmitz, S.J. de Vries, A. Bordogna, L. Bonati,
P.L. Kastritis and A.M.J.J. Bonvin. Defining the limits of homology modelling in
information-driven protein docking Proteins: Struc. Funct. \& Bioinformatics,
81, 2119-2128 (2013).}

\item {\bf G.C.P. van Zundert}, S. Jaeqx, G. Berden, J.M. Bakker, K.
Kleinermanns, J. Oomens, A.M. Rijs. IR spectroscopy of isolated neutral and
protonated adenine and 9-methyladenine. ChemPhysChem 12, 1921--1927 (2011).

\stopitemize

\stopBackmatterHead

\stopcomponent
