\startcomponent chapter6

\environment layout
\product thesis

\define\CA{C\low{\m{\alpha}}}
\define\boldP{{\bf P}}
\define\boldC{{\bf C}}

\Chapter{Inferring interface residues from distance-restraints for
high-resolution HADDOCKing}


\Section{Introduction}


In this Chapter, we introduce a method to infer residues that are likely to be
at the interface when distance-restraints are available in addition to
high-resolution atomic models of the chains. The method is benchmarked on 125
complexes taken from the protein-protein docking benchmark 4.0 when 3, 5 and 7
restraints are available with an upper distance of 30\Angstrom, caparable to
the information content that is provided by DSS/BS3 mass spectrometry/crosslink
data. Finally, we show how the results can be combined with HADDOCK to
complement the docking runs with regular unambiguous distance-restraints.


\Section{Methods}


\Subsection{Inferring interface residues from distance-restraints}

In the previous Chapter, we have introduced the concept of the accessible
interaction space, the set of all possible complexes that are consistent with
a certain number of distance-restraints. Indeed, the presence of distance-restraints between
two interacting macromolecular biomolecules can significantly reduce the
accessible interaction space. To infer residues that are likely to be at the
interface, we assume that these residues are often found to be interacting in
the whole interaction space. By performing a full-exhaustive 6 dimensional
search of the three translational and three rotational degrees of freedom and
counting the number of interactions that each solvent accessible residue forms
in complexes consistent with a certain number of restraints, important residues
can be determined. We define two residues to be interacting when the \CA--\
\CA\ distance is smaller than 10\Angstrom. We only consider the \CA-atoms of
solvent accessible residues of both chains to make the computations more
tractable. The average number of interactions per complex that a residue \m{i}
forms is given by

\placeformula[eq:count-interactions]
\startformula
\overline{N}_{\text{inter}}^i = \frac{\sum_R^{\bf P} w_R \sum_C^{\boldC_R} I_{C}}
{\sum_R^\boldP w_R \boldC_R}
\stopformula

where the first summation is over all rotations \m{\boldP} indexed by \m{R};
\m{w_R} is a weight factor to unbias certain rotations/orientations; the second
summation is over all complexes \m{\boldC_R} that are formed within a
translational scan indexed by \m{c}; and \m{I_{C}} is the number of
interactions that are formed by residue \m{i} in each sampled complex.

This approach has been implemented in DisVis, which requires for the
interaction analysis an extra input file containing the residue numbers for the
fixed and scanning chain that are solvent accessible. DisVis then outputs a
file containing the number of interactions that are formed by each residue for
complexes consistent with at least \m{N} restraints.


\Subsection{Benchmarking interface residue extraction}

We benchmarked our approach on 125 complexes taken from the protein-protein
docking benchmark 4.0. The complexes were taken such that at least 7 virtual
cross-links were available, determined by the XWalk software.  The virtual
crosslinks were picked such that the solvent accessible surface (SAS) distance
was shorter or equal than 34\Angstrom, and the Euclidean distance smaller or
equal than 30\Angstrom.  The cross-links were randomly picked from the list
using the probability distribution as was used by Kahraman: 0 --
10\Angstrom.... to mimic cross-link data.

The solvent accessible residues were determined by running naccess on the two
unbound proteins. Residues that had a relative solvent accessiblity of the main
or side chain of 50 were used as surface residues. DisVis runs were performed
for each complex using 3, 5, and 7 random restraints with a 5.27\Deg\
rotational sampling, and default values for the voxelspacing voxelspacing
(1\Angstrom), and maximum clashing volume (200\Angstrom\high3) and minimum
interaction volume (300\Angstrom\high3). The average interactions per complex
were calculated for complexes consistent with all restraints. 

Correct interface residues were taken from the complex structure using the
above definition of interaction, under the restriction that the residue was
also present in the unbound proteins. To analyse the success, the precision and
recall were calculated, given by

\placeformula[eq:precision]
\startformula
\text{P} = \frac{\text{TP}}{\text{TP} + \text{FP}}
\stopformula

\placeformula[eq:recall]
\startformula
\text{R} = \frac{\text{TP}}{\text{TP} + \text{FN}}
\stopformula

where TP, FP and FN stand for True Positive, False Positive, and False
Negative, respectively.


\Subsection{HADDOCKing with distance-restraints}






\Section{Results}


\Subsection{Residues with many interactions are more likely to be at the interface}


\Subsection{HADDOCKing with distance-restraints}


\Section{Conclusions}

\stopcomponent
